---
title: "DATA474_FinalProj"
author: "Ava Mizrahi"
date: "2026-05-05"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
```


```{r}
spotify <- read.csv("spotify.csv")
dim(spotify)
glimpse(spotify)
summary(spotify)
```


```{r}
# cleaning
spotify <- spotify %>%
  distinct %>%
  mutate(
    explicit = as.factor(explicit),
    track_genre = as.factor(track_genre),
    duration_min = duration_ms / 60000 )

spotify <- spotify %>%
  mutate(
    track_genre = case_when(
      track_genre %in% c("latin", "latino") ~ "latin_latino",
      track_genre %in% c("reggae", "reggaeton") ~ "reggae_reggaeton",
      track_genre %in% c("dub", "dubstep") ~ "dub_dubstep",
      track_genre %in% c("singer-songwriter", "songwriter") ~ "singer_songwriter",
      track_genre %in% c("punk", "punk-rock") ~ "punk",
      track_genre %in% c("indie", "indie-pop") ~ "indie",
      track_genre %in% c("minimal-techno", "techno") ~ "techno",
      track_genre %in% c("pop", "pop-film") ~ "pop",
      TRUE ~ as.character(track_genre)))

spotify$track_genre <- as.factor(spotify$track_genre)

```

```{r}
summary(spotify)
```





```{r}
library(nnet)

set.seed(1)

train_index <- sample(1:nrow(spotify), 0.7 * nrow(spotify))

train_data <- spotify[train_index, ]
test_data <- spotify[-train_index, ]
genre_model <- multinom(
  track_genre ~ popularity + explicit + danceability + energy + key + mode + speechiness + acousticness + instrumentalness + liveness + valence + tempo + time_signature + duration_min, data = train_data, MaxNWts = 5000, maxit = 500)
```


```{r}
# Grouping top 5 popular grenres
spotify %>%
  group_by(track_genre) %>%
  summarise(avg_pop = mean(popularity, na.rm = TRUE)) %>%
  slice_max(order_by = avg_pop, n=5)

# Grouping middle 5 popular genres
spotify %>%
  group_by(track_genre) %>%
  summarise(median_pop = median(popularity, na.rm = TRUE)) %>%
  arrange(desc(median_pop)) %>%
  mutate(rank = row_number(), total = n()) %>%
  filter(rank >= (total/2 - 2) & rank <= (total/2 + 2))

# Grouping least 5 popular genres
spotify %>%
  group_by(track_genre) %>%
  summarise(avg_pop = mean(popularity, na.rm = TRUE)) %>%
  arrange(avg_pop) %>%
  head(5)
```


```{r}

odds_table <- as.data.frame(exp(coef(genre_model)))

odds_table$genre <- rownames(odds_table)

odds_table <- odds_table %>%
  select(genre, everything())

odds_long <- odds_table %>%
  pivot_longer(cols = -genre, names_to = "characteristic", values_to = "odds_ratio") %>%
  filter(characteristic != "(Intercept)") %>%
  mutate(effect_strength = abs(log(odds_ratio)), direction = ifelse(odds_ratio > 1, "increases odds", "decreases odds"))
```


```{r}
# Top 5 genres odds ratios
selected_genres_top <- c("k-pop", "chill", "pop", "sad", "grunge")

selected_odds <- odds_table %>%
  filter(genre %in% selected_genres_top)

selected_odds
```


```{r}
# Middle 5 genres odds ratios
selected_genres_middle <- c("german", "garage", "children", "ska", "cantopop")

selected_odds <- odds_table %>%
  filter(genre %in% selected_genres_middle)

selected_odds
```

```{r}
# Least 5 genres odds ratios
selected_genres_low <- c("iranian", "romance", "classical", "detroit-techno", "chicago-house")
selected_odds <- odds_table %>%
  filter(genre %in% selected_genres_low)

selected_odds
```


```{r}
# Which characteristic have the best predictability
odds_long %>%
  group_by(characteristic) %>%
  summarise(avg_strength = mean(effect_strength)) %>%
  arrange(desc(avg_strength))
```



```{r}
# pseudo R-squared
null_model <- multinom(track_genre ~ 1, data = train_data, trace = FALSE)

1 - (deviance(genre_model) / deviance(null_model))
```



```{r}
library(caret)

# predictions
pred_genre <- predict(genre_model, newdata = test_data)

# confusion matrix
cm <- confusionMatrix(data = pred_genre, reference = test_data$track_genre)
cm
```



```{r}
# precision and F1
metrics <- as.data.frame(cm$byClass)

metrics_summary <- metrics %>%
  summarise(avg_precision = mean(Precision, na.rm = TRUE), avg_f1 = mean(F1, na.rm = TRUE))

metrics_summary
```

Model performance:
Accuracy: 20.6%
Precision	17.5%
F1-score	18.7%


Conclusions:
- model has 20.6% accuracy, which is better than randomly guessing (.94%)
- pseudo-R^2 was 0.281, meaning the model improved over a baseline model with no predictors
- some genres are easier to classify (distinctive genres were predicted better because they have more unique audio features)
- similar genres get confused because they overlap a lot




```{r}
# 20 most popular genres
top_20_popular <- spotify %>%
  group_by(track_genre) %>%
  summarise(avg_popularity = mean(popularity, na.rm = TRUE)) %>%
  arrange(desc(avg_popularity)) %>%
  slice_head(n = 20) %>%
  pull(track_genre)

spotify_top20 <- spotify %>%
  filter(track_genre %in% top_20_popular) %>%
  droplevels()
```

```{r}
library(nnet)

set.seed(1)

train_index <- sample(1:nrow(spotify_top20), 0.7 * nrow(spotify_top20))

train_data <- spotify_top20[train_index, ]
test_data <- spotify_top20[-train_index, ]

genre_model_top20 <- multinom(
  track_genre ~ popularity + explicit + danceability + energy + key + mode + speechiness + acousticness + instrumentalness + liveness + valence + tempo + time_signature + duration_min,
  data = train_data,
  MaxNWts = 5000,
  maxit = 500
)
```


```{r}
# Predictability 
pred_top20 <- predict(genre_model_top20, newdata = test_data)

mean(pred_top20 == test_data$track_genre)
```

```{r}
# Odds ratios
odds_table_top20 <- as.data.frame(exp(coef(genre_model_top20)))

odds_table_top20$genre <- rownames(odds_table_top20)

odds_table_top20 <- odds_table_top20 %>%
  select(genre, everything())

odds_long_top20 <- odds_table_top20 %>%
  pivot_longer(cols = -genre, names_to = "characteristic", values_to = "odds_ratio") %>%
  filter(characteristic != "(Intercept)") %>%
  mutate(effect_strength = abs(log(odds_ratio)))

odds_long_top20 %>%
  group_by(characteristic) %>%
  summarise(avg_strength = mean(effect_strength)) %>%
  arrange(desc(avg_strength))
```

```{r}
# Psuedo R squared
null_model_top20 <- multinom(track_genre ~ 1, data = train_data, trace = FALSE)

1 - (deviance(genre_model_top20) / deviance(null_model_top20))
```

```{r}
# Confusion Matrix
library(caret)

pred_top20 <- predict(genre_model_top20, newdata = test_data)

cm_top20 <- confusionMatrix(data = pred_top20, reference = test_data$track_genre)

cm_top20
```

```{r}
# Precision and F1
metrics_top20 <- as.data.frame(cm_top20$byClass)

metrics_top20 %>%
  summarise(
    avg_precision = mean(Precision, na.rm = TRUE),
    avg_f1 = mean(F1, na.rm = TRUE)
  )
```

Top 20 model performance:
Accuracy: 33.2%
Precision: 32.2%
F1: 30.5%
Pseduo R^2: 0.323
Predicted well: metalcore, pagode, ambient, mandopop, sertanejo 
predicted poorly: anime, indian, k-pop, electronic 


```{r}
# Comparing top 20 model to full model
model_comparison_table <- tibble(
  Model = c("All Genres Model", "Top 20 Genres Model"), Accuracy = c(20.6, 33.2), Precision = c(17.5, 32.2), F1_Score = c(18.7, 30.5), Pseudo_R2 = c(28.1, 32.3))

model_comparison
```





