---
title: "DATA_Final_Proj_LogisticReg"
author: "Ava Mizrahi"
date: "2026-05-09"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


```{r}
spotify <- read.csv("spotify.csv")
dim(spotify)
glimpse(spotify)
summary(spotify)
```

```{r}
# cleaning
spotify <- spotify %>%
  distinct() %>%
  mutate(
    explicit = as.factor(explicit),
    track_genre = as.factor(track_genre),
    duration_min = duration_ms / 60000 )

spotify <- spotify %>%
  mutate(
    track_genre = case_when(
      track_genre %in% c("metal", "metalcore") ~ "metal",
      track_genre %in% c("deep-house", "electronic", "progressive-house") ~ "electronic_house",
      track_genre %in% c("minimal-techno", "techno") ~ "techno",
      track_genre %in% c("pop", "pop-film") ~ "pop",
      TRUE ~ as.character(track_genre)))

# 20 most popular genres
top_20_popular <- spotify %>%
  group_by(track_genre) %>%
  summarise(avg_popularity = mean(popularity, na.rm = TRUE)) %>%
  arrange(desc(avg_popularity)) %>%
  slice_head(n = 20) %>%
  pull(track_genre)

spotify_top20 <- spotify %>%
  filter(track_genre %in% top_20_popular) %>%
  mutate(track_genre = factor(track_genre))

```

```{r}
summary(spotify)
```

```{r}
# checking for distinctness in genres 
top20_summary <- spotify_top20 %>%
  group_by(track_genre) %>%
  summarise(
    danceability = mean(danceability),
    energy = mean(energy),
    acousticness = mean(acousticness),
    speechiness = mean(speechiness),
    instrumentalness = mean(instrumentalness),
    valence = mean(valence),
    tempo = mean(tempo)
  )

top20_summary
```



```{r}
# multinomial logistic regression
library(nnet)

set.seed(1)

train_index <- sample(1:nrow(spotify_top20), 0.7 * nrow(spotify_top20))

train_data <- spotify_top20[train_index, ]
test_data <- spotify_top20[-train_index, ]

genre_model_top20 <- multinom(
  track_genre ~ popularity + explicit + danceability + energy + key + mode + speechiness + acousticness + instrumentalness + liveness + valence + tempo + time_signature + duration_min,
  data = train_data,
  MaxNWts = 5000,
  maxit = 500
)
```


```{r}
# Predictability 
pred_top20 <- predict(genre_model_top20, newdata = test_data)

mean(pred_top20 == test_data$track_genre)
```


```{r}
# Odds ratios
odds_table_top20 <- as.data.frame(exp(coef(genre_model_top20)))

odds_table_top20$genre <- rownames(odds_table_top20)

odds_table_top20 <- odds_table_top20 %>%
  select(genre, everything())

odds_long_top20 <- odds_table_top20 %>%
  pivot_longer(cols = -genre, names_to = "characteristic", values_to = "odds_ratio") %>%
  filter(characteristic != "(Intercept)") %>%
  mutate(effect_strength = abs(log(odds_ratio)))

odds_long_top20 %>%
  group_by(characteristic) %>%
  summarise(avg_strength = mean(effect_strength)) %>%
  arrange(desc(avg_strength))
```

```{r}
# Pseudo R squared
null_model_top20 <- multinom(track_genre ~ 1, data = train_data, trace = FALSE)

1 - (deviance(genre_model_top20) / deviance(null_model_top20))
```

```{r}
# Confusion Matrix
library(caret)

pred_top20 <- factor(pred_top20, levels = levels(test_data$track_genre))

cm_top20 <- confusionMatrix(
  data = pred_top20,
  reference = test_data$track_genre
)
cm_top20
```

```{r}
# Precision and F1
metrics_top20 <- as.data.frame(cm_top20$byClass)

metrics_top20 %>%
  summarise(avg_precision = mean(Precision, na.rm = TRUE), avg_f1 = mean(F1, na.rm = TRUE))
```


This model used multinomial logistic regression to classify spotify songs into the 20 genres with the highest average popularity (24,000 songs). Similar genres were combined prior to modeling to reduce overlap. Instrumentalness, speechiness, energy, danceability, and acousticness are the top 5 strongest predictors of genre. Audio texture and style features are more useful for genre classification than musical attributes like key, tempo, or popularity. 

The model achieved:
- 35.69% accuracy
- 29.9976% precision
- 27.82% F1 Score
- .308 pseudo R squared

Genres such as metal, electronic house, ambient, and sertanejo were easier to classify because they have more distinct audio features. Broader genres like pop were more difficult to distinguish. 










